---
title: Congress network analysis
---

In this exercise we will use graphs in R and look at some graph summary metrics
to better understand collaboration networks in the US House. We have data from the
117th Congress (the current congress) on cosponsorship of bills, and we will use
this data to define a graph or network where representatives are the nodes and
cosponsorship of bills are the edges.

The dataset we have lists the sponsor and cosponsor(s) of every bill introduced in the
117th House where this data is available from the Government Printing Office. We will
create a cosponsorship network where we create an edge between every cosponsor and
the sponsor of that bill (we are not creating links between cosponsors of the same bill,
as they may not actually have worked together). We are only including original cosponsors
of the bill, not ones that signed on later.

First, as always, we load libraries. We are using tidyverse and a graph
manipulation library (igraph) that we haven't used yet.

```{r}
library(tidyverse)
library(igraph)
```

Next, we read the data.

```{r}
data = read_csv(here::here("networks", "sponsors.csv"))
head(data)
```

To turn this into a graph, we need to format it into a data frame that has columns
for the start and end of every edge, but we don't want to have duplicate edges.
First, we need to modify the data frame to only include a single record for each sponsor/cosponsor
pair. This is a little bit tricky because we are using an undirected graph - we don't
want duplicate records for the same two representatives when one was a sponsor and
the other was a cosponsor, and vice-versa. To do this, we create a new data frame
with sponsor1 and sponsor2 columns, and just put whoever comes earlier in alphabetical
order in the sponsor1 column.

```{r}
sponsors = mutate(data,
                  sponsor1=if_else(sponsor < cosponsor, sponsor, cosponsor),
                  sponsor2=if_else(sponsor < cosponsor, cosponsor, sponsor)
                  ) |> select(-c(sponsor, cosponsor))
head(sponsors)
```

Now we just need to create a version of the data with a single entry for each
sponsor pair.

```{r}
sponsor_pairs = sponsors |> group_by(sponsor1, sponsor2) |> summarize() |> ungroup()
```

Finally, we can create the graph.

```{r}
graph = graph_from_data_frame(sponsor_pairs, directed=F)
```

## Shortest paths

Just like we did with road networks, we can compute shortest paths in the Congress network.
Representative Ralph Norman [R-SC] is considered the most conservative house member, and
Barbara Lee [D-CA] is the most liberal
https://www.govtrack.us/congress/members/report-cards/2020/house/ideology
Let's compute the shortest path between them

```{r}
shortest_paths(graph, "Rep. Lee, Barbara [D-CA]", "Rep. Norman, Ralph [R-SC]")
```

That's a bit surprising - I would not have expected to only need one intermediary
(Rep. Andy Biggs [R-AZ]) to get from the most liberal to most conservative member of
the House. But it's true; we can look at the data to confirm.

First, find bills that were sponsored by Lee and cosponsored by Biggs, or vice versa
Biggs will always be sponsor1 since his name is first alphabetically

```{r}
# answer
sponsors |>
    filter(sponsor1=="Rep. Biggs, Andy [R-AZ]", sponsor2=="Rep. Lee, Barbara [D-CA]")
```

Lee introduced HR 256 and Biggs cosponsored, which was a repeal of the 2002 Congressional
authorization of use of force in Iraq (https://www.congress.gov/bill/117th-congress/house-bill/256/cosponsors).

Biggs will also appear first in the bill cosponsored with Norman

```{r}
# answer
sponsors |>
    filter(sponsor1=="Rep. Biggs, Andy [R-AZ]", sponsor2=="Rep. Norman, Ralph [R-SC]")
```

These representatives have cosponsored many bills, perhaps less surprising since
they are both Republicans, and Biggs is fairly conservative.

That said, it's still somewhat surprising that the most conservative and most liberal members
of Congress are so close together in cosponsorship space. Maybe this is just a fluke because
Lee sponsored a popular bill? Let's try with two other very liberal and conservative members,
Alexandria Ocasio-Cortez [D-NY] and Madison Cawthorn [R-NC]. Make sure you spell them exactly the way
they are in the data (starting with Rep. and then with last name first, and finally with party and state
in square brackets).

```{r}
# answer
shortest_paths(graph, "Rep. Ocasio-Cortez, Alexandria [D-NY]", "Rep. Cawthorn, Madison [R-NC]")
```

No, this doesn't appear to be a fluke. We can use some graph-level metrics to determine how prevalent
this is. One such metric is the mean distance - how long is the average path between any
two representatives?

```{r}
mean_distance(graph)
```

That is pretty short.

It's worth checking how many graph "components" there are - a component is a
disconnected part of the graph with no paths to the other representatives. If there
is only one component, then all nodes in the graph are directly or indirectly connected to
all others. If there are many compomnents, the mean distance could be artificially low
because it's not including distances between representatives in different components.

```{r}
count_components(graph)
```

There is only one component; every representative is connected to every other representative.

We can compute all of the distances also, for all pairs.

```{r}
dists = distances(graph)
```

What is longest distance?

```{r}
max(dists)
```

What is the distribution of distances?

The table function just tabulates the total number of observations for each value.

```{r}
table(dists)
```

let's find the pairs with the longest distance between them - you could consider
these representatives the "most dissimilar."
The "which" function will find the indices where the expression is true, and
the arr.ind tells it to return original array/matrix indices.

```{r}
which(dists == 4, arr.ind=T)
```

Another measure of connectedness is the edge density, which is simply a ratio of
how many connections there are in the graph compared to how many there could be.

This is pretty easy to calculate

```{r}
total_edges = length(E(graph))
total_nodes = length(V(graph))

# How many possible edges are there in the graph?
possible_edges = total_nodes * (total_nodes - 1) / 2

# now we can calculate edge density
total_edges / possible_edges
```

igraph has a built in function for edge density as well, which you should use (it's easier and less error prone to use pre-written functions than write your own).

By default, this function assumes that "loop edges" from a node to itself don't exist,
which they don't in this graph, but if you did have loop edges you would need to
specify loop=T to correctly calculate the denominator for this metric.

```{r}
edge_density(graph)
```

## Centrality metrics

### Degree centrality

There are a bunch of different types of centrality metrics, all of which aim to
measure how central a given node (or edge, but we won't look at that here) is
to the graph overall. The simplest is "degree centrality", which is just the "degree"
or number of edges connected to a node. Does the degree centrality suggest any reason
why Mike Carey and Cedric Richmond were so far apart?

```{r}
degree(graph) |> sort()
```

### Closeness centrality

More complex types of centrality include closeness centrality, which is the inverse
of the total distance from a node to all others. So higher closeness centrality indicates
you can reach more representatives with fewer hops. Closeness centrality is often
normalized by multiplying by the number of nodes - 1 (https://en.wikipedia.org/wiki/Closeness_centrality)
so it represents the inverse of the average distance rather than the sum.

```{r}
closeness(graph, normalize=T) |> sort()
```

### Betweenness centrality

Another type of centrality metric is betweenness centrality - this measures how
often a representative is on the shortest path between all pairs of other representativess.
It could be treated as a measure of the power a representative has to connect others,
though treating it as an "importance" can be tricky. People often assume that representatives
with high betweeness centrality are very important to the network and if removed would increase
the distance significantly, but this is not necessarily the case as betweenness centrality
doesn't mean there is not another representative that could play a similar role.

```{r}
betweenness(graph) |> sort()
```

The final metric of centrality we'll consider is eigenvector centrality. The
centrality of each node is a scaled sum of the centralities of all the nodes it's
connected to, so a representative that is connected to other representatives that are highly
connected scores more highly than a representative that is connected to the same number of
less connected representatives. As it happens, this is also the eigenvector of the adjacency
matrix (i.e. a square matrix with the nodes as rows and columns and a 1 if there is an edge between
the nodes referenced by the rows and columns). This is roughly the algorithm used by
Google to rank web pages, or at least it was when Google first started.

Unlike other centrality metrics, the eigen_centrality function returns a list. We can
get the vector of centralities with $vector

```{r}
eigen_centrality(graph)$vector |> sort()
```

We can put all of the centrality metrics together into a single table.

```{r}
centralities = tibble(name=names(degree(graph)), degree=degree(graph), betweenness=betweenness(graph), closeness=closeness(graph), eigen=eigen_centrality(graph)$vector)
select(centralities, -name) |> cor()
centralities |> arrange(closeness) |> View()
```

Do the different centrality metrics tell similar stories? If they differ, why do you think they do?

## Weighted graphs

We've been working with an unweighted graph up until now, meaning the edges were all the same. However,
it's reasonable to think that people who cosponsor more often might be more connected. We can use a
weighted graph instead. In a weighted graph, each edge is assigned a "weight," which can be thought of
as a length or distance. Then a shortest path algorithm will not find the fewest number of links, but the
lowest total weight. For instance, we can create a graph where the weight is the reciprocal of the number
of times two representatives worked together. All we need to do is create a weight column in the
data frame we use to create the graph. Copy the graph creation code from above and modify it to have
a weight column that is the reciprocal of the number of times two representatives worked together. Recalulate the
centrality metrics

```{r}
# answer
sponsor_pairs = group_by(sponsors, sponsor1, sponsor2) |>
  summarize(weight=1/n()) |>
  ungroup()

weight_graph = graph_from_data_frame(sponsor_pairs, directed=F)

centralities = tibble(name=names(degree(weight_graph)), degree=degree(weight_graph), betweenness=betweenness(weight_graph), closeness=closeness(weight_graph), eigen=eigen_centrality(weight_graph)$vector)

View(centralities)
```

How do the centralities compare to before?

Some representatives have become much more important in some centrality measures
(e.g. Eleanor Holmes Norton [D-DC], who is a delegate rather than a representative
because residents of the District of Columbia do not have a vote in Congress.)
Why do you think this is?

Some representatives have become more central because they co-sponsor a lot of bills,
some of which may be co-sponsored by many representatives. Rebuild the graph and repeat the
analysis using only bills with 5 or fewer cosponsors. How do the results change?

```{r}
# answer
sponsor_pairs = sponsors |>
    group_by(bill) |>
    filter(n() <= 5) |>
    group_by(sponsor1, sponsor2) |>
    summarize(weight=1/n())

weight_graph = graph_from_data_frame(sponsor_pairs, directed=F)

centralities = tibble(name=names(degree(weight_graph)), degree=degree(weight_graph), betweenness=betweenness(weight_graph), closeness=closeness(weight_graph), eigen=eigen_centrality(weight_graph)$vector)

View(centralities)
```