---
title: Chapel Hill Flooding
author: Matt Bhagat-Conway
---

## Load libraries

As always, we start by loading libraries. Here, we see a new library: the `sf` library, designed for working with spatial data in R.

```{r}
library(tidyverse)
library(sf) # this is the "simple features" package, designed for GIS work in R
library(ggspatial) # this provides some extra features for mapping with ggplot
library(leaflet)
```

## Reading in data

Up to now, we've used `read_csv` to read CSV files and `read_excel` to read Excel files. In all cases, these files were quite small and not sensitive, and could be stored in the git repository itself. The spatial datasets we're working with here are much larger, too large to be stored in the repository itself. Instead, we will have to download them from Canvas, and store them someplace else on our computers.

The problem is that this will be a different place on everyone's computer. For example, I might have the files at `/Users/mwbc/unc/gis-data/`, but you will have them in a different location. We could just have everyone adjust the path themselves, but if we were working collaboratively on a project that would quickly get problematic. Instead, we will use _environment variables_ and an `.Renviron` file to specify where the files are on each machine. Environment variables are just text names and values that can be shared across multiple R scripts (and, depending on how you set them up, other programs as well). This way, if you run your code on a different computer, you just have to update `.Renviron` and all the code will work.

Environment variables are defined outside of your R script or Quarto document. There are many ways to set up environment variables, but we are going to use an `.Renviron` file. First, download the GIS data from Canvas and unzip it, and put the folder somewhere on your computer.

Next, we will create an environment variable to tell R where to find the files. Create a `.Renviron` file by choosing File -> New File -> Text File. By convention, the names of environment variables are all caps with words separated by underscores, so we will call our environment variable `DATA_PATH`. Put this in your `.Renviron`:

DATA_PATH="/path/to/plan372_gis_data"

On Windows, the path will start with `C:\`. If there are backslashes (i.e. `\`, not `/`) in the path, you need to replace each single backslash with a double backslash.

Save the file as `.Renviron` in your `plan372-exercises` folder (without the backticks). Make sure the file is called `.Renviron`, not `.Renviron.txt`.

We can access environment variables in R using the `Sys.getenv` function. However, first we need to restart R (which we will have to do any time we change `.Renviron`). Do this by choosing Session -> Restart R, and then run the cell above to load libraries.

If you have set up your `.Renviron` correctly, the cell below should store the environment variable `DATA_PATH` in the R variable `DATA_PATH`, and then print out the contents of the GIS data directory.

```{r}
DATA_PATH = Sys.getenv("DATA_PATH")
list.files(DATA_PATH)
```

You should get this output:

    [1] "chapel_hill_transit_stops.csv"  "goraleigh-stops.txt"            "orange_durham_osm_highways.dbf"
    [4] "orange_durham_osm_highways.prj" "orange_durham_osm_highways.shp" "orange_durham_osm_highways.shx"
    [7] "orange_durham_parcels.cpg"      "orange_durham_parcels.dbf"      "orange_durham_parcels.prj"     
    [10] "orange_durham_parcels.shp"      "orange_durham_parcels.shx"      "orange_durham_parcels.xml"     
    [13] "restaurant_inspections_geo.csv" "wake_osm_highways.dbf"          "wake_osm_highways.prj"         
    [16] "wake_osm_highways.shp"          "wake_osm_highways.shx"          "Waterbodies.cpg"               
    [19] "Waterbodies.dbf"                "Waterbodies.prj"                "Waterbodies.shp"               
    [22] "Waterbodies.shx" 

If you don't, your path is not set correctly in your `.Renviron`. If you see `plan372_gis_data` in the output, you just need to add `/plan372_gis_data` to the end of the path in `.Renviron`, and restart R.

## Working with GIS data

This script works through how to work with spatial data using the following datasets
 
1. orange_durham_osm_highways: Roads in Orange and Durham counties, from OpenStreetMap
2. Waterbodies: Bodies of water in the Chapel Hill area, from Chapel Hill Open Data
3. orange_durham_parcels: Parcels (plots of land) in Orange and Durham counties
4. chapel_hill_transit_stops.csv: bus stups in chapel hill, contains number of the average number of buses per day for each stop

### Reading data

First, we read in our datasets. To read a GIS file, we use the `st_read` function, which comes from `sf`. `st_read` can read most vector GIS datasets. A vector GIS dataset just means a dataset of points, lines, or polygons, with associated attributes. Let's start with highways. This is a shapefile (note the .shp extension)

`file.path` just puts two (or more) paths together. So `file.path(DATA_PATH, "orange_durham_osm_highways.shp")` just refers to the `orange_durham_osm_highways.shp` file inside the folder specified by DATA_PATH.

```{r}
orange_durham_osm_highways = st_read(file.path(DATA_PATH, "orange_durham_osm_highways.shp"))

#as always, good practice is to inspect your data. Is everything looking as expected?
head(orange_durham_osm_highways)
```

Notice the geometry field at the end. This is what tells R where/how to map the object.

### Making a map

Our plan is to make a map of Chapel Hill/Durham's highways/road system and plot bus stops. We have already read in the highways. We also need to read in the water bodies. Do that now.

```{r}
# answer:
waterbodies = read_sf(file.path(DATA_PATH, "waterbodies.shp"))
```

There are a number of packages for making maps. The easiest is to just use one we've already been using: `ggplot`. This will map the roads and water bodies in Chapel Hill.

```{r}
ggplot() +
  layer_spatial(data = waterbodies) +
  layer_spatial(data = orange_durham_osm_highways)
```

That is pretty ugly. We can zoom in on Chapel Hill. The units of `xmin` and `xmax` are degrees of longitude, and `ymin` and `ymax` are degrees of latitude, respectively. You can find the degrees of latitude and longitude for any location in many ways, but the simplest is just to right-click on a location on Google Maps. Here we use `annotation_spatial` instead of `layer_spatial` to keep the map from auto-zooming to the full extents of our data. `crs=4326` specifies that we are using degrees of latitude and longitude (more on that soon).

```{r}
ggplot() +
  annotation_spatial(data = waterbodies) +
  annotation_spatial(data = orange_durham_osm_highways) +
  shadow_spatial(st_bbox(c(xmin=-79.0755, xmax=-79.0031, ymin=35.8943, ymax=35.9511), crs=4326))
```

That is looking much better, but we might want the water bodies to be in blue.

```{r}
ggplot() +
  annotation_spatial(data = waterbodies, color = "blue", fill = "blue") +
  annotation_spatial(data = orange_durham_osm_highways) +
  shadow_spatial(st_bbox(c(xmin=-79.0755, xmax=-79.0031, ymin=35.8943, ymax=35.9511), crs=4326)) +
  # adding theme_minimal will get rid of the gray background
  theme_minimal() +
  # and this will remove the coordinates from the axes
  theme(
    axis.text = element_blank(),
    panel.grid = element_blank()
  )
```

This is a little different than how we've used ggplot in the past - we specified the data inside the geoms themselves because we're using two different datasets, and we haven't specified an aes because there are no data attributes being mapped onto plot characteristics.

#### Legends

With a simple map like this, it's pretty easy to tell what is what - the black is obviously roads/paths and the blue is obviously water. It's good practice to include a legend anyhow, and it will be important on more complex maps.

`ggplot` only creates legends for values specified in an aes. We can specify an aes with the label of each layer, rather than a column (note that it must be in quotes).

```{r}
ggplot() +
  annotation_spatial(data = waterbodies, aes(color="Water", fill="Water")) +
  annotation_spatial(data = orange_durham_osm_highways, aes(color="Roadways")) +
  shadow_spatial(st_bbox(c(xmin=-79.0755, xmax=-79.0031, ymin=35.8943, ymax=35.9511), crs=4326)) +
  # adding theme_minimal will get rid of the gray background
  theme_minimal() +
  # and this will remove the coordinates from the axes
  theme(
    axis.text = element_blank(),
    panel.grid = element_blank()
  )
```

That looks... pretty bad. We need to tell ggplot how we want each part of our map colored. We do this with `scale_fill_manual` and `scale_color_manual` to tell it what color "Water" and "Roadways" are.

```{r}
ggplot() +
  annotation_spatial(data = waterbodies, aes(color="Water", fill="Water")) +
  annotation_spatial(data = orange_durham_osm_highways, aes(color="Roadways")) +
  shadow_spatial(st_bbox(c(xmin=-79.0755, xmax=-79.0031, ymin=35.8943, ymax=35.9511), crs=4326)) +
  # adding theme_minimal will get rid of the gray background
  theme_minimal() +
  # and this will remove the coordinates from the axes and the grid from the map
  theme(
    axis.text = element_blank(),
    panel.grid = element_blank()
  ) +
  scale_color_manual(values=c("Water"="blue", "Roadways"="gray"), aesthetics=c("color", "fill"))
  
```

This is still non-ideal; we really want to combine those. We can tell ggplot to combine those into a single legend, titled "Legend", by telling ggplot to make legends for the color and fill and call them both "Legend":

```{r}
ggplot() +
  annotation_spatial(data = waterbodies, aes(color="Water", fill="Water")) +
  annotation_spatial(data = orange_durham_osm_highways, aes(color="Roadways")) +
  shadow_spatial(st_bbox(c(xmin=-79.0755, xmax=-79.0031, ymin=35.8943, ymax=35.9511), crs=4326)) +
  # adding theme_minimal will get rid of the gray background
  theme_minimal() +
  theme(
    axis.text = element_blank(),
    panel.grid = element_blank()
  ) +
  scale_color_manual(values=c("Water"="blue", "Roadways"="gray"), aesthetics=c("color", "fill")) +
  guides(color=guide_legend(title="Legend"), fill=guide_legend(title="Legend"))
```

#### Adding CSV data to the map

There is another file in the GIS data folder, `chapel_hill_transit_stops.csv` which contains information on the location of transit stops in Chapel Hill. Read in that CSV file, and call the dataset `bus_stops`.

```{r}
# answer:
bus_stops = read_csv(file.path(DATA_PATH, "chapel_hill_transit_stops.csv"))
```

If we try to just add that layer to the map directly, it won't work:

```{r}
#| error: true
ggplot() +
  annotation_spatial(data = bus_stops, aes(color = "Bus Stop")) +
  annotation_spatial(data = waterbodies, aes(color="Water", fill="Water")) +
  annotation_spatial(data = orange_durham_osm_highways, aes(color="Roadways")) +
  shadow_spatial(st_bbox(c(xmin=-79.0755, xmax=-79.0031, ymin=35.8943, ymax=35.9511), crs=4326)) +
  # adding theme_minimal will get rid of the gray background
  theme_minimal() +
  theme(
    axis.text = element_blank(),
    panel.grid = element_blank()
  ) +
  scale_color_manual(values=c("Water"="blue", "Roadways"="gray", "Bus Stop"="red"), aesthetics=c("color", "fill")) +
  guides(color=guide_legend(title="Legend"), fill=guide_legend(title="Legend"))
```

The bus stops came from a CSV file, not a GIS file, so they don't automatically have geometry associated with them. Take a look at the data and see if there are any columns we might use to figure out where these stops are:

```{r}
head(bus_stops)
```

It looks like columns `stop_lat` and `stop_lon` have the latitude and longitude values of the stops. We just need to tell R about what the columns are. `crs=4326` just tells R to interpret them as latitude/longitude (we'll discuss more about this in the next exercise). We tell R that the coordinates are `stop_lon` and `stop_lat` (while that order may seem backwards, GIS usually puts east-west location first and north-south second).

```{r}
bus_stops = st_as_sf(bus_stops, coords=c("stop_lon", "stop_lat"), crs=4326)
```

Now we can try that map again:

```{r}
ggplot() +
  annotation_spatial(data = bus_stops, aes(color = "Bus Stop")) +
  annotation_spatial(data = waterbodies, aes(color="Water", fill="Water")) +
  annotation_spatial(data = orange_durham_osm_highways, aes(color="Roadways")) +
  shadow_spatial(st_bbox(c(xmin=-79.0755, xmax=-79.0031, ymin=35.8943, ymax=35.9511), crs=4326)) +
  # adding theme_minimal will get rid of the gray background
  theme_minimal() +
  theme(
    axis.text = element_blank(),
    panel.grid = element_blank()
  ) +
  scale_color_manual(values=c("Water"="blue", "Roadways"="gray", "Bus Stop"="red"), aesthetics=c("color", "fill")) +
  guides(color=guide_legend(title="Legend"), fill=guide_legend(title="Legend"))
```

This is better, but the bus stops are appearing behind the roads. We can fix this by just changing the order of the layers; layers are drawn bottom-to-top in the order they are specified.

```{r}
ggplot() +
  annotation_spatial(data = waterbodies, aes(color="Water", fill="Water")) +
  annotation_spatial(data = orange_durham_osm_highways, aes(color="Roadways")) +
  annotation_spatial(data = bus_stops, aes(color = "Bus Stop")) +
  shadow_spatial(st_bbox(c(xmin=-79.0755, xmax=-79.0031, ymin=35.8943, ymax=35.9511), crs=4326)) +
  # adding theme_minimal will get rid of the gray background
  theme_minimal() +
  theme(
    axis.text = element_blank(),
    panel.grid = element_blank()
  ) +
  scale_color_manual(values=c("Water"="blue", "Roadways"="gray", "Bus Stop"="red"), aesthetics=c("color", "fill")) +
  guides(color=guide_legend(title="Legend"), fill=guide_legend(title="Legend"))
```

#### Mapping data values to map aesthetics

There's also some extra data in the bus stops dataset which shows how many buses per day serve each stop. We can map those to map attributes just like we did for any other type of plot in ggplot:

```{r}
ggplot() +
  annotation_spatial(data = waterbodies, color="Blue", fill="Blue") +
  annotation_spatial(data = orange_durham_osm_highways, color="Gray") +
  annotation_spatial(data = bus_stops, aes(color = buses_per_day)) +
  shadow_spatial(st_bbox(c(xmin=-79.0755, xmax=-79.0031, ymin=35.8943, ymax=35.9511), crs=4326)) +  # adding theme_minimal will get rid of the gray background
  theme_minimal() +
  theme(
    axis.text = element_blank(),
    panel.grid = element_blank()
  )
```

I've also simplified the legend a bit by removing the legend entries for water and roads as these are basically just background information. If you wanted to keep them, you would need to be able to specify color aesthetics and scales for them separately from the bus stops. ggplot on its own can't do this, but if you wanted to do this the `ggnewscale` package can.

It's hard to make much sense of that map; almost all the bus stops appear black. That's probably because of outliers; the busiest stop looks like it serves 500 buses a day! Make a histogram of buses per day to check.

```{r}
# answer:
bus_stops |>
  ggplot(aes(x=buses_per_day)) +
    geom_histogram()
```

Indeed there are a few high outliers. To make the map more readable, we can apply a different color scheme. It is typical on maps to use a binned color scheme where a single color represents a range; it is easier for folks to figure out what the value for a particular location might be.

To do this, you have to tell ggplot what you want the breakpoints between different colors to be. Here, I am setting them to quintiles of the distribution of buses per day, so there are roughly equal numbers of bus stops in each category.

```{r}
ggplot() +
  annotation_spatial(data = waterbodies, color="Blue", fill="Blue") +
  annotation_spatial(data = orange_durham_osm_highways, color="Gray") +
  annotation_spatial(data = bus_stops, aes(color = buses_per_day)) +
  shadow_spatial(st_bbox(c(xmin=-79.0755, xmax=-79.0031, ymin=35.8943, ymax=35.9511), crs=4326)) +  # adding theme_minimal will get rid of the gray background
  # adding theme_minimal will get rid of the gray background
  theme_minimal() +
  theme(
    axis.text = element_blank(),
    panel.grid = element_blank()
  ) +
  scale_color_binned(
    # unname here just makes the legend be labeled in terms of the actual
    # values rather than 20%, 40%, 60%, etc.
    # this is also one of the few places in ggplot where you have to use
    # $ notation for columns.
    breaks=unname(quantile(bus_stops$buses_per_day, c(0, 0.2, 0.4, 0.6, 0.8, 1))),
    limits=range(bus_stops$buses_per_day),
    palette="Blues"
  )
```

## Other map elements

In addition to a legend, maps often have a scale bar to help folks interpret what they're looking at, and attribution for data sources. Those are both easy to add. We add an `annotation_scale()` from the ggspatial package to get a scale bar.

```{r}
ggplot() +
  annotation_spatial(data = waterbodies, color="Blue", fill="Blue") +
  annotation_spatial(data = orange_durham_osm_highways, color="Gray") +
  annotation_spatial(data = bus_stops, aes(color = buses_per_day)) +
  shadow_spatial(st_bbox(c(xmin=-79.0755, xmax=-79.0031, ymin=35.8943, ymax=35.9511), crs=4326)) +  # adding theme_minimal will get rid of the gray background
  # adding theme_minimal will get rid of the gray background
  theme_minimal() +
  theme(
    axis.text = element_blank(),
    panel.grid = element_blank()
  ) +
  scale_color_binned(
    # unname here just makes the legend be labeled in terms of the actual
    # values rather than 20%, 40%, 60%, etc.
    # this is also one of the few places in ggplot where you have to use
    # $ notation for columns.
    breaks=unname(quantile(bus_stops$buses_per_day, c(0, 0.2, 0.4, 0.6, 0.8, 1))),
    limits=range(bus_stops$buses_per_day),
    palette="Blues"
  ) +
  annotation_scale()
```

It is also typical to add attribution for any data you got from elsewhere in the lower right (and in some cases this may be a legal requirement). We can do this with the ggplot `annotate` function. Setting the position to `Inf, -Inf` just says to place the text as far right and as far down as possible, and hjust/vjust adjust the positioning to bring the text back onto the map.

```{r}
ggplot() +
  annotation_spatial(data = waterbodies, color="Blue", fill="Blue") +
  annotation_spatial(data = orange_durham_osm_highways, color="Gray") +
  annotation_spatial(data = bus_stops, aes(color = buses_per_day)) +
  shadow_spatial(st_bbox(c(xmin=-79.0755, xmax=-79.0031, ymin=35.8943, ymax=35.9511), crs=4326)) +  # adding theme_minimal will get rid of the gray background
  # adding theme_minimal will get rid of the gray background
  theme_minimal() +
  theme(
    axis.text = element_blank(),
    panel.grid = element_blank()
  ) +
  scale_color_binned(
    # unname here just makes the legend be labeled in terms of the actual
    # values rather than 20%, 40%, 60%, etc.
    # this is also one of the few places in ggplot where you have to use
    # $ notation for columns.
    breaks=unname(quantile(bus_stops$buses_per_day, c(0, 0.2, 0.4, 0.6, 0.8, 1))),
    limits=range(bus_stops$buses_per_day),
    palette="Blues"
  ) +
  annotation_scale() +
  annotate("label", Inf, -Inf, label="Map data © OpenStreetMap contributors, Town of Chapel Hill", fill="white", hjust=1.01, vjust=-0.1)
```

## Quick data exploration with Leaflet

ggplot is great for high-quality maps, but sometimes we just want something easy and interactive to explore our data. The leaflet package provides that. We can add our bus stops to an interactive map like this:

```{r}
leaflet() |>
  addTiles() |>
  addCircleMarkers(data=bus_stops, popup=~stop_name)
```

In addition to circles for point locations, you can also draw lines. To keep this from being too slow, I am also telling it to filter the roadways just to major roads (using the [OpenStreetMap roadway taxonomy](https://wiki.openstreetmap.org/wiki/Key:highway)).

```{r}
leaflet() |>
  addTiles() |>
  addPolylines(data=filter(
      orange_durham_osm_highways,
      highway %in% c("motorway", "trunk", "primary", "secondary", "tertiary")
    ), popup=~name) |>
    addCircleMarkers(data=bus_stops, popup=~stop_name)

```