---
title: Chapel Hill flooding
---

Sometimes you may want to combine the best of both worlds: do some of your analysis in QGIS, then export to R for additional analysis (or vice-verse).

We can export data from QGIS as CSV and analyze it using the same tools we've used in R in the past.

As always, first, we need to load libraries.


```{r}
library(tidyverse)

```

Next, we read in our data file. You may need to adjust the path.


```{r}
data = read_csv("chapel_hill_flooding_fromqgis.csv")

```


```{r}
data$PIN[duplicated(data$PIN)]

```

it's probably a good idea to go back to QGIS to see if these are simply duplicates,
or something else.

it looks like these aren't duplicates, but parcels that are split into multiple pieces e.g. by a road
The permits dataset, though, has records for the entire parcel. We'll get an incorrect
answer if we use this, because we'll be essentially duplicating the development that occurred
for each part of a parcel. So we should group the parcels and sum up the areas, but not sum
up the development costs, so we have one row per parcel that accurately reflects development


```{r}
parcels = group_by(data, PIN) |>
  summarize(
    Shapearea=sum(Shapearea),  # we want to sum up the areas of the parts of the parcel
    parcel_permits_total_construction_cost=first(parcel_permits_total_construction_cost),

    # if any part of the parcel is in the flood area, we want to call it a flood area
    floodarea=any(floodarea==1)
  )

```

Use the filter function to remove non-Chapel Hill parcels by filtering
for non-NA construction cost. In this case, the parcel_permits data file contained
records for all of Chapel Hill, even if parcels didn't have any permits pulled, and
just had a zero for the constrction cost.

If you were doing a similar analysis, you'd want to check this - you wouldn't
want to accidentally exclude parcels that hadn't had any permits pulled.


```{r}
# answer

```

Now, use group_by and summarize to compute development intensity inside and
outside the flood zone. Think about whether you are computing development intensity
per square foot, or per square foot per lot

I am summing up total cost and total area by flood zone / non flood zone, and
then computing intensity afterwards. If I just took a mean of the already-calculated
development intensity, I would undercount large parcels, because they would contribute
the same to the mean as small parcels, whereas in actuality development intensity large parcels 
represents more development.


```{r}
# answer

```



