---
title: Linear regression
---

Linear regression is a tool for understanding relationships between multiple variables, and predicting
outcomes.

We will use data on home sales in Charlotte to build a model of home sales price.
We have data on 500 single-family home sales in Mecklenburg County from 2010 to
the present. This includes the sale price of the home as well as a number of
attributes of the home collected by the tax assessor.

First, we need to load libraries.

```{r}
library(tidyverse)
library(ggthemes)

```

Next, we'll load our data


```{r}
data = read_csv(here::here("linear-regression", "data", "charlotte_home_sales.csv"))
head(data)

```

## Exploratory analysis

The columns of the data are
- dte_dateof - date of sale
- houseno, houseunit, stdir, stname, sttype, stsuffix - components of street address
  (number, unit number, street direction, street name, street type, and street suffix)
- amt_price - Sale price in dollars
- heatedarea - area of the home that is heated (i.e. living area), square feet
- yearbuilt - year the home was built
- fullbaths - number of full baths
- halfbaths - number of half baths
- bedrooms - number of bedrooms
- actype - type of air conditioning/cooling
- vacantorim - VACant or IMProved (there is a building on the lot)
- totalac - total acreage of the lot
- siding - type of siding (wall coverage)
- area - the part of the Charlotte metropolitan area where the home is located
  (Public Use Microdata Area code)


Before we start doing any regressions, it's always a good idea to get to know our data
a bit first. Make a histogram of sale price:


```{r}
# answer

```

Make a histogram of heated area


```{r}
# answer

```


Now, we can start thinking about building our model. One variable that is likely
to be associated with home prices is the size of the home (heatedarea). We can
make a scatterplot of these variables before we start thinking about a model.


```{r}
ggplot(data, aes(x=heatedarea, y=amt_price)) +
  geom_point(size=0.1)

```

It's hard to read because of the outliers (we'll return to this issue shortly)
For now, we can just set the limits to look at the bulk of the data.

```{r}
ggplot(data, aes(x=heatedarea, y=amt_price)) +
  geom_point(size=0.1) +
  xlim(0, 5000) +
  ylim(0, 2000000)

```

## Regression models with one variable

It does look like there's a positive trend. We're ready to build our first
model! We'll express price as a function of heated area In R, you use the lm()
function to estimate a linear regression. The first argument is a _formula_,
which expresses the mathematical equation you want to estimate. The dependent
variable is on the left, followed by a ~, followed by the independent
variable(s). We also specify the dataset we are using for estimation. We do
not need to specify a constant/intercept. R will include that automatically.


```{r}
single_variable_model = lm(amt_price~heatedarea, data)

# to see the results of our model, we can run summary()
summary(single_variable_model)

```

The first section includes some information about the residuals - the
difference between the predictions and the actual values in the data. The next
section details the coefficients - what the estimate was, and how much random
variation there might be in that estimate due to sampling error (we'll cover
this in a moment). The last interesting statistic is the R-squared, which
we'll discuss below.

## Seeing our model visually

Let's plot our regression line on top of our scatterplot. We can use
geom_abline for this, which accepts an intercept and slope. The intercept is
the coefficient for the constant. In a regression with only one variable, the
coefficient for that variable is the slope of the line that predicts the
outcome. Hint: e+05 is scientific notation - 1.5e+05 would be 1.5
times 10 to the 5th, or 150,000.


```{r}
ggplot(data, aes(x=heatedarea, y=amt_price)) +
  geom_point(size=0.1) +
  geom_abline(intercept=-1.175e5, slope=206.2, color="red") +
  xlim(0, 5000) +
  ylim(0, 2000000)

```

## Dealing with outliers

That line doesn't look like it goes quite through the middle of the data. If
we remove the limits on the plot we can see why:


```{r}
ggplot(data, aes(x=heatedarea, y=amt_price)) +
  geom_point(size=0.1) +
  geom_abline(intercept=-98341, slope=196.4, color="red")

```

there are outliers well off to the right and top. Because regression is
minimizing the sum of _squared_ residuals, outliers can have a very large
influence. Even though there are not very many of them, they are far from the
regression line. We can remove the outliers and run the model again. Whether
or not to remove outliers is up to you and what makes sense for your analysis.
However, you should avoid "cherry-picking" data - filtering the data until you
get the results you want from your analysis.

Exercise: filter the data to only homes with heated area less than 5,000
square feet, and prices less than $2 million. Save the result in a new variable
`no_outliers`. Run a model on this filtered dataset.


```{r}
# answer

```

Make a similar plot.


```{r}
ggplot(no_outliers, aes(x=heatedarea, y=amt_price)) +
  geom_point(size=0.1) +
  geom_abline(intercept=-2254.106, slope=148.3, color="red")

```

Are the results different?


## Multiple regression

So far we've only included a single variable in the model - heated area. But
with regression, we can include many variables. Then the coefficient for each
independent variable will be how much change in the dependent variable is
associated with a 1-unit change in the independent variable, holding
everything else in the model constant.

For example, these data were collected over the course of 10+ years. During
that time, housing prices went up significantly. We'd expect homes that sold
in 2010 to sell for less than homes that sold in 2020. We can add a variable
for year of sale, but first we need to compute the year of sale.

```{r}
no_outliers = no_outliers |> mutate(year_of_sale=year(dte_dateof))

year_built_model = lm(amt_price~heatedarea+year_of_sale, no_outliers)
summary(year_built_model)

```

What is the effect of year of sale?

When two variables are correlated, including them both in the model can do
strange things. As an example, let's try adding number of bedrooms to our
model


```{r}
bedrooms_model = lm(amt_price~heatedarea+year_of_sale+bedrooms, no_outliers)
summary(bedrooms_model)

```

What is the relationship between an additional bedroom and price?
Is it what you would expect?

Exercise: Estimate the same model again, but without the heated area. Does the
result change?

```{r}
# answer

```

Is the coefficient for bedrooms now what you would expect?

What's going on? In the first model, we estimated the relationship between
number of bedrooms and price holding square footage constant - so the value of
an additional bedroom without increasing the total size of the home. The trend
lately has been towards larger rooms, and adding a room without increasing the
size of the home means smaller rooms, so it is reasonable to think that an
additional bedroom in a home of the same size might not be particularly
valuable. The model above shows a slightly negative coefficient, though it's
not _statistically significant_ (we'll discuss that in a bit), so we can't be
sure it wouldn't be positive if we had a different sample of 1500 homes.

Here, we plot the relationship between heated area and price, with color
indicating number of bedrooms. Note that the colors at the bottom and top of
the point cloud are all mixed together, indicating that there isn't really a
trend in price for different numbers of bedrooms with the same square footage.


```{r}
ggplot(no_outliers, aes(x=heatedarea, y=amt_price, color=as.factor(bedrooms))) +
  geom_point() +
  ylim(0, 1000000) +
  scale_color_colorblind()

```


This is known as Simpson's Paradox - cases where adding a variable can change
the sign of the coefficient of another variable. It is a specific case of the
more general omitted variable bias and collinearity concepts. Omitted variable
bias is the notion that whenever you add a variable to a model, it may change
the coefficients for any other related variable. In general, this means that
you can only interpret the coefficients of a regression in relation to what
else is included in the model. For instance, you would likely find that the
number of TVs a household owns is highly predictive of the price of their
home. But this is not because more TVs cause households to buy more expensive
homes - rather, it is likely that higher incomes lead to buying more expensive
homes _and_ to buying more TVs. Adding income into the model would likely make
the effect of TVs go away - because then the model would be asking the
question of how an additional TV is related to home value, holding income
constant - and it's probably not very related.

In extreme cases, adding additional variables to the model that are correlated
with each other may result in counterintuitive, statistically-insignificant
results. This is know as a collinearity or multicollinearity problem, and is
something you should look into if you are experiencing unexpected results in a
regression.

## Categorical variables

So far we've only included continuous, numerical variables in the model. But
we know other things likely affect home prices as well, for instance what the
home looks like from the street. The data contain a variable "siding" that
says what the siding (wall covering) on the outside of the home is. People may
pay more for certain types of siding if they are more durable, more
attractive, etc. But it's not a number, so we can't just put it in the model
with a coefficient.

The most common way to put a categorical variable into a regression model is
to use "dummy variables" There will be one variable for each category, that is
one if the home is in that category, and zero otherwise. This allows the model
to estimate a coefficient for each of the categories—so for instance we will
have the amount of money brick siding is associated with, or vinyl siding,
etc. You have to leave one category out; that is the base category, and
everything else is relative to that. (If we didn't, there wouldn't be a unique
set of coefficients for the computer to estimate. You could get exactly the
same prediction from many models, by increasing the coefficient for the
constant and decreasing the coefficient for all of the dummy variables.)

We can create dummy variables in R by including non-numeric values in the
regression. R will automatically create one variable for each category. If you
wanted to create dummy variables for a numeric column (e.g., if siding types
were coded as 1, 2, 3, etc.), you would convert the column to a factor using
as.factor() and then include that converted column in your model.


```{r}
siding_model = lm(amt_price~heatedarea+year_of_sale+siding, no_outliers)
summary(siding_model)

```

What category did R choose as the base category?

How much more valuable is Hardiplank than Vinyl?

It is often said that the three most important aspects of a home are location,
location, and location. Similar homes in different neighborhoods may sell for
very different prices. There is an area variable in the file, which represents
different parts of the Charlotte metro area. Exercise: add this as a
categorical variable.

```{r}
# answer

```

How much do the different areas differ in price?

You may also see coefficients for dummy variables for areas or groups referred
to as "fixed effects".

## Evaluating the quality of the model

Let's look at our model with heated area, year of sale, and siding type The
first thing to look at is the R-squared. This tells us how much of the
variation in price is explained by the variables we're using - in this case,
63.4% of the variation is explained. The Adjusted R2 adjusts the R2 based on
how many independent variables you are using - if we used enough independent
variables, we could fit the data really well, but not be able to predict well.
Adjusted R2 reduces the R2 a bit based on how many variables are used.

Next is the statistical significance of the coefficients. Every coefficient is
estimated with some error, because we're using a finite dataset - we only have
1,500 observations, and far more properties have sold recently, and even more
could sell in the future. The standard error quantifies how much variation
there might be if we ran the regression on a different sample.

## Hypothesis tests of regression coefficients

We found a coefficient of 147 for heated area, and a standard error of 4.192.
We want to know whether this value is _statistically significant_, i.e. far
enough from zero that we can be fairly confident we didn't find it just due to
the particular sample of properties we used. To do this, we use a hypothesis
test. If you're familiar with hypothesis tests of means, this works the same
way, but if you're not we'll go over it from the beginning.

With a hypothesis test, you have a null hypothesis and an alternate
hypothesis. The null hypothesis in regression is that the true coefficient
value is 0 - i.e. there is no relationship between heated area and price. We
test how likely we would be to find a relationship this large, given our data
and sample size, if the null hypothesis were true. If it is sufficiently
unlikely, we conclude that the relationship is likely real, and not due to
sampling error.

The way we test this is by dividing the coefficient by the standard error -
this is the number of standard errors away from zero the coefficient is. This
is known as a t-value. Then, we use a distribution to determine the
probability of of getting a result this many standard deviations from 0 if the
true relationship were zero.

This requires assuming a distribution for the coefficients. A distribution is
just a graph that indicates what the probability of observing particular
t-values is, if the null hypothesis were true. The most common distribution is
the normal distribution or "bell curve", which I plot below.


```{r}
t = seq(-4, 4, 0.05)
y = dnorm(t)
plot_data = tibble(t=t, y=y)
ggplot(plot_data, aes(x=t, y=y)) +
  geom_line()

```

The higher the line is, the more likely it is that we would get a coefficient
of that t-value if the null hypothesis were true.

While the normal distribution is common throughout statistics, in a
regression, the distribution we use is actually the t-distribution. This
distribution accounts for possible inaccuracy in the estimation of the
standard deviation. This difference only really matters when the sample size
of your regression is very small; once the number of observations gets to a a
few dozen, the t and normal distributions are almost identical.

R computes the hypothesis tests for your linear regression automatically, but
we'll work through computing one by hand, just for experience. This code
computes the t-value

```{r}
tval = 147 / 4.192

```

This computes the probability of getting a value larger than 147 if the
null hypothesis is true. This is using a Student's t distribution; hypothesis
testing in a linear regression uses this distribution, like a comparison of
means would if you've taken intro stats. When your sample size is large
(hundreds or thousands more observations than independent variables) the t
distribution is almost identical to the normal distribution.


```{r}
prob_larger = 1 - pt(tval, nrow(no_outliers) - 2)

```

This is the probability of getting a value larger than the one we got, if the
null hypothesis were true. But that means we are implicitly assuming that the
coefficient will always be positive. What we really want is to know the
probability of getting a coefficient _this far from zero_ - so we also want to
add the probability of getting a coefficient less than -147 with this standard
error. Since the t and normal distributions are symmetrical, we can just
multiply the value we got above by 2. This is called a "two-tailed test" in
statistics.

```{r}
prob_larger = prob_larger * 2
prob_larger

```

The probability is zero. Of course, it isn't exactly zero (anything could
happen randomly), but it's close enough to zero that the computer can't tell
the difference. We can be pretty certain that the null hypothesis that the
coefficient is zero is not true. Generally, probabilities below 0.05 are
considered "statistically significant," but there is some debate about this.
This probability is known as a p-value

To reiterate what is meant by this probability: this is the probability of
getting a coefficient this extreme if the null hypothesis were true. It is
_not_ the probability that the null hypothesis is true, or the probability
that your model is correct.

You generally won't do statistical tests manually like this. In the model
output, R prints these tests automatically. The t-values are in the third
column, and the p-values in the fourth. Heated area is presented as <2e-16 -
which indicates it is smaller than the smallest value the computer can
represent.

```{r}
summary(siding_model)

```

Are there any variables that are not statistically significant in this model?
What about in your model that included area?

## Predicting from a linear regression

A linear regression is just estimating an
equation that predicts the dependent variable based on the independent
variables. We can write out the equation. For instance, suppose we wanted to
estimate the price of a 2000 square foot home sold in 2021, with Hardiplank
siding.

```{r}
-4.309e7 + 1.437e2 * 2000 + 2.134e4 * 2021 + 9.881e4

```

Exercise: use your model that includes area to predict the price of a 2000
square foot home sold in 2021 with Masonite siding in area 03103.


```{r}
# answer

```

Normally, you wouldn't make predictions manually like this. Instead, you would
create a new dataset that had all of your independent variables, and then
apply the predict() function to it. For instance, here I'll make a dataset of
several different types of homes, and predict the sales prices of all of them.
A developer might use a model like this to forecast profits from a new
development---probably a bit more complex model with more home features. Here
we create a table with the candidate homes. Each line is a column in the
table. Lines where there is only one item will have that item repreated for
every row of the table. We use the tribble function, which lets us specify a
table directly in R code.


```{r}
candidate_homes = tribble(
    ~heatedarea, ~year_of_sale, ~siding,
    1200,        2021,          "HARDIPLANK",
    1500,        2021,          "HARDIPLANK",
    2000,        2021,          "HARDIPLANK",
    1200,        2021,          "FACE BRICK",
    1500,        2021,          "FACE BRICK",
    2000,        2021,          "FACE BRICK",
)

```

now, we can use predict() to create a new column in candidate_homes with the
forecasted price. It should be close to what we found above (a little different
because we used rounded coefficients above but predict() uses the original,
un-rounded coefficients).


```{r}
candidate_homes$forecast_price = predict(siding_model, candidate_homes)
candidate_homes

```

Exercise: Estimate the price of a 2000 square foot home sold in 2021 with
Hardiplank siding in areas 03103, 03105, or 03108 (note: you will need to put
the area codes in quotes). A developer might use a result like this to decide
which of several sites would be most profitable to build on. You will need to use the
model that included area for your predictions to be meaningful.


```{r}
# answer

```

While prediction is often seen as a primary function of linear regression, it
is actually not that common in urban analytics. People use linear models for
interpretation more often - to interpret the coefficients and see what the
relationships between different variables are, holding other things constant.

These models had quite high predictive power (R2), which is common in models
of home prices. In models of other phenomena (especially behavioral
phenomena), much lower R2 is common. This may represent an issue for
prediction, but does not necessarily represent an issue for interpretation.
It's only an issue for interpretation if there are other variables left out
that would affect the coefficients for the variables included.

Regression has a lot more pieces than we discussed here. I recommend the books
An Introduction to Statistical Learning with Applications in R, and The
Effect, if you want to learn more. Both are available online as free ebooks.
