---
title: Missouri dealership sales data
---

The State of Missouri releases sales data from car dealerships at
https://dor.mo.gov/public-reports/dealers_franchise_report.txt
Download this file, and we'll work through reading it into R. It's in a non-standard
format intended to be read by humans, but we can use regular expressions to read
it into a format we can analyze.


```{r}
library(tidyverse)

```

First, we will read the text file into a single string. We are using read_file here
which will read the file as a single string rather than a table like read_csv


```{r}
str = read_file(here::here("text_processing", "dealers_franchise_report.txt"))

```

Now, we need to split the file into a list/vector, with one item for each dealership
we can do this with the str_split function, which takes the string, and a regular
expression that identifies where we want it to be split. We want to write a regular
expression that will match the DEALER****** delimiter between dealers. Since * is a special
character in regular expressions, we need to _escape_ it by places two backslashes (\\)
before it. This tells R to use it to match a *, not to modify the character before it

A list in R is str_split returns data as a list of lists - with one list for the components of each input
string. Since we only had one input string, we will retrieve the first list. [[]] in R is a _scalar indexing_ operator; it retrieves a single value at the _index_ specified. In R, indices start from 1.


```{r}
dealers = str_split(str, "DEALER# \\**")[[1]]

```

This will show us the first three records:


```{r}
dealers[1:3]

```

The first item in the list of dealerships is the header for the whole file. We can
discard that.

The `c` function creates a _vector_—i.e. a group of values that are all of the same type. By indexing by a negative vector, we remove those records.


```{r}
dealers = dealers[-c(1)]

```

The dealership records are kind of a mess, because all the lines have been smooshed
together. We can split each dealership record into lines, again with str_split. This
time, we don't use [1], because we want a list of lists - a list whose elements are
lists of all the lines associated with each dealership.
\n is the code in R and most other programming languages for a newline (return)


```{r}
dealers = str_split(dealers, "\n")

```

We want to extract the following information for each dealership:
- Dealer number
- Dealership name
- Dealership phone number
- New, used, and total cars ("units") sold in each year

The table should contain the columns dealer_id, dealer_name, dealer_phone, year, new_units, used_units

We will start by parsing the information for a single dealership, as this will be easier to work with.
Then, we'll use that code to process all of the dealerships.

### Get the first dealership


```{r}
dealer = dealers[[1]]
dealer

```

First, we want to find the first line, which contains the dealer name, dealer id, and phone number.


```{r}
dealer_info_line = first(dealer[str_detect(dealer, "^[:whitespace:]*[:alpha:]{1,2}[:digit:]+.*PHONE: [:digit:]{3}-[:digit:]{3}-[:digit:]{4}")])

```

Now, we'll use a regular expression that will extract the dealer number, the dealer name, and the phone number from the dealer_info line, and store it in the dealer_info variable. Use the str_match function.


```{r}
dealer_info = str_match(dealer_info_line, "^[:whitespace:]*([:alpha:]{1,2}[:digit:]+)[:whitespace:]*(.*)[:whitespace:]*PHONE: ([:digit:]{3}-[:digit:]{3}-[:digit:]{4})")

```

now, we need to extract the per-year sales from the remaining lines
We need to write a regular expression that will detect the lines, and then use (a variation of)
it to extract the lines with years. Store the output in the count_lines variable.


```{r}
count_lines = dealer[str_detect(dealer, "^[:whitespace:]*UNITS SOLD IN [:digit:]{4}[:whitespace:]+NEW:[:whitespace:]*[:digit:]+[:whitespace:]+USED:[:whitespace:]*[:digit:]+[:whitespace:]+TOTAL:[:whitespace:]*[:digit:]+[:whitespace:]*$")]

```

Now, use str_match to extract the relevant information from the count lines


```{r}
counts = str_match(count_lines, "^[:whitespace:]*UNITS SOLD IN ([:digit:]{4})[:whitespace:]+NEW:[:whitespace:]*([:digit:]+)[:whitespace:]+USED:[:whitespace:]*([:digit:]+)[:whitespace:]+TOTAL:[:whitespace:]*([:digit:]+)[:whitespace:]*$")

counts

```

Give the counts column names


```{r}
colnames(counts) = c("all", "year", "new", "used", "total")

```

Next, we convert the counts to a tibble (tabular data).


```{r}
counts = as_tibble(counts)

```

We now have a table (counts) with all of the information about the counts. We just need
to add the dealer name, number, and phone to it to have our final dataset.


```{r}
counts$dealer_id = dealer_info[1, 2]
counts$dealer_name = dealer_info[1, 3]
counts$dealer_phone = dealer_info[1, 4]

```

and we can get rid of the "all" column with the full matched line


```{r}
counts = select(counts, -all)

```

Now, we want to do this for every dealership in the data. To do this, we will
write a "function." We've been using functions since we started with R - a function
is stored R code that is executed with certain arguments, and returns some value.
For instance, read_csv is a function that takes the names of a file as the argument,
and returns the data read from the file.

Here, we create our own function to process an individual dealer record, and use map
to apply it to each dealer records. We then use the function list_rbind (list row bind)
to convert the rows returned each time the function was called into a tibble.


```{r}
dealer_table = map(dealers, function (dealer) {
    # first, we want to find the first line, which contains the dealer name, dealer id, and phone number.
    dealer_info_line = first(dealer[str_detect(dealer, "^[:whitespace:]*[:alpha:]{1,2}[:digit:]+.*PHONE: [:digit:]{3}-[:digit:]{3}-[:digit:]{4}")])
  
    # now, write a regular expression that will extract the dealer number, the dealer name, and the phone
    # number from the dealer_info line
    dealer_info = str_match(dealer_info_line, "^[:whitespace:]*([:alpha:]{1,2}[:digit:]+)[:whitespace:]*(.*)[:whitespace:]*PHONE: ([:digit:]{3}-[:digit:]{3}-[:digit:]{4})")
    
    # now, we need to extract the per-year sales from the remaining lines
    # We need to write a regular expression that will detect the lines, and then use (a variation of)
    # it to extract the lines with years
    count_lines = dealer[str_detect(dealer, "^[:whitespace:]*UNITS SOLD IN [:digit:]{4}[:whitespace:]+NEW:[:whitespace:]*[:digit:]+[:whitespace:]+USED:[:whitespace:]*[:digit:]+[:whitespace:]+TOTAL:[:whitespace:]*[:digit:]+[:whitespace:]*$")]
    
    # now, extract the relevant information
    counts = str_match(count_lines, "^[:whitespace:]*UNITS SOLD IN ([:digit:]{4})[:whitespace:]+NEW:[:whitespace:]*([:digit:]+)[:whitespace:]+USED:[:whitespace:]*([:digit:]+)[:whitespace:]+TOTAL:[:whitespace:]*([:digit:]+)[:whitespace:]*$")
    
    counts
    
    # give the counts column names
    colnames(counts) = c("all", "year", "new", "used", "total")
    # convert the counts to a tibble (tabular data)
    counts = as_tibble(counts)
    
    # we now have a table (counts) with all of the information about the counts. We just need
    # to add the dealer name, number, and phone to it to have our final dataset
    counts$dealer_id = dealer_info[1, 2]
    counts$dealer_name = dealer_info[1, 3]
    counts$dealer_phone = dealer_info[1, 4]
    
    counts = select(counts, -all)
    
    return(counts)
  }) |>
  list_rbind()

head(dealer_table)

```

## Using regular expressions for data cleaning

The dealer names likely include a lot of whitespace at the end of them (depending on how exactly
you wrote your regular expression), and some end with INC, ", INC", "LLC", "LLC." or some variation
of this (with or without comma after name, with or without period). We want to clean up these names
in order to have the names in our dataset match the names people use to refer to these dealerships.

We can use the str_replace function to replace a regular expression with a value. Write a regular expression
to match the whitespace and any LLC/INC at the end of the dealer name. To match either LLC or INC, you can
use "alternation" - two patterns in parentheses with a pipe symbol | (above the enter key) between them
will match one or the other of the patterns. Recall also that two backslashes before a special character
such as . will match that character, rather than whatever the special character would normally match.
The empty string at the end of str_replace indicates to replace with nothing.


```{r}
# answer

```

### Exercise: plot the number of used, new, and total cars sold over time in the state of Missouri


```{r}
# answer

```

### Exercise: what dealership sold the most cars in 2022?


```{r}
dealer_table |>
    filter(year==2022) |>
    group_by(dealer_id) |>
    summarize(total=sum(total), dealer_name=first(dealer_name)) |>
    arrange(-total)

```
